---
title: "02_Missing"
author: "Margaret Anderson"
date: "2026-06-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries
```{r}
library(fpp3)
library(dplyr)
library(lubridate)
library(tidyr)
library(EpiReport)
library(ggplot2)
library(broom)
library(lmtest)
library(ggfortify)
library(car)
library(forecast)
library(readxl)
library(reshape2)
library(openxlsx)
library(magrittr)
library(skimr)
library(pacman)
library(stringr)
library(naniar)
library(rio)
library(here)
library(flextable)
library(officer)
select <- dplyr::select 
```

Read in cleaned dataset for 2014 thru 2025
```{r}
surveillance_doi <- readRDS(here("5_RMD", "surveillance_doi.rds"))
```

STEP 1: Overall assessment of missingness patterns prior to remove of placeholder districts and mid-year activations for missingness analysis with no removal or assessment of placeholder districts - using the surveillance doi dataset
```{r}
surveillance_doi |>
  group_by(AN, Week, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  ggplot(aes(x = Week, y = AN, fill = PctMissing)) +
  geom_tile() +
  facet_wrap(~ Disease) +
  scale_fill_gradient(low = "white", high = "red", name = "Missing %") +
  scale_y_continuous(breaks = 2014:2025, expand = c(0, 0)) +
  labs(
    title = "S#: Missingness Heatmap by Week and Year Per Disease of Interest",
    x = "Week",
    y = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.text = element_text(color = "black", size = 8),
    legend.title = element_text(color = "black", size = 8),
    plot.title = element_text(color = "black", size = 8),
    axis.title = element_text(color = "black", size = 8),
    axis.text.y = element_text(color = "black", size = 6),
    axis.text.x = element_text(color = "black", size = 8),
    strip.text = element_text(color = "black", size = 8),
    strip.background = element_blank()
  )

ggsave(here("6_visuals", "S1_missing_heatmapNPOPLACEHOLDERREMOVAL.pdf"), width = 10, height = 8, dpi = 300)
```
Step 2: Diagnosing districts with either 100 or near 100% missing, these were used as placeholders by the surveillance team, here we identify those. There are also mid-year activations of districts, we use >40% missing to identify those 
```{r}
#check for missing by disease, district, and year for those over 94% (placeholders)
surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  group_by(AN, DISTRICT, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  filter(PctMissing >= 94) |>
  arrange(AN, DISTRICT, Disease) |> 
  print(n = Inf)

#check for missing by disease, district, and year for those over 40% (mid year activations)
surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  group_by(AN, DISTRICT, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  filter(PctMissing >= 40) |>
  arrange(AN, DISTRICT, Disease) |> 
  print(n = Inf)

#finding from surveillance_doi is that we are not 100% missing in some diseases, we need to diagnose this, it is only the case in 2015
surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  filter(DISTRICT %in% c("Djamane Mbarissou", "Gagal", "Koyom", "Lagon", "Torrock")) |>
  filter(AN == 2015) |>
  group_by(DISTRICT, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    TotalCases = sum(Cases, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(DISTRICT, Disease)|> 
  print(n = Inf)

surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  filter(DISTRICT == "Bekourou") |>
  filter(AN == 2015) |>
  select(DISTRICT, AN, Week, Disease, Cases) |>
  arrange(Disease, Week)|> 
  print(n = Inf)

surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  filter(DISTRICT == "Bekourou") |>
  group_by(DISTRICT, AN, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    TotalCases = sum(Cases, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(PctMissing >= 80) |>
  arrange(Disease, AN, DISTRICT)|> 
  print(n = Inf)

#conclusion from Bekourou diagnostic is that Bekourou was a placeholder 
surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  filter(DISTRICT %in% c("Djamane Mbarissou", "Gagal", "Koyom", "Lagon", "Torrock")) |>
  filter(AN == 2015) |>
  select(DISTRICT, AN, Week, Disease, Cases) |>
  arrange(DISTRICT, Disease, Week)

#Check for Bekourou in 2014 - conclusion: not in 2014
surveillance_doi |>
  filter(DISTRICT == "Bekourou") |>
  distinct(AN) |>
  arrange(AN)|> 
  print(n = Inf)

#check what years we have for the 5 placeholder districts 
surveillance_doi |>
  filter(DISTRICT %in% c("Djamane Mbarissou", "Gagal", "Koyom", "Lagon", "Torrock")) |>
  distinct(DISTRICT, AN) |>
  arrange(DISTRICT, AN)|> 
  print(n = Inf)

surveillance_doi |> summarise(min(AN), max(AN))


```

STEP 3: Create Figures for Supplemental Material to show the mid-year missingness patterns for the 2016 mid-year activation districts. Excluded guinea worm due to high missing and complementary surveillance program via Carter Center 

# ============================================================
# Figure S#: Weekly Cases by Disease — 2016 Mid-Year Activation Districts
# Supplemental Figure for methods transparency
# ============================================================

```{r}
##Diagnose mid-year exclusions 
surveillance_doi |>
  filter(Disease != "Guinea Worm") |>
  filter(!(DISTRICT %in% c("Djamane Mbarissou", "Gagal", "Koyom", "Lagon", "Torrock") & AN == 2015)) |>
  filter(!(DISTRICT == "Bekourou" & AN %in% c(2015, 2016, 2017))) |>
  group_by(DISTRICT, AN, Disease) |>
  summarise(
    FirstNonNA_Week = min(Week[!is.na(Cases)], na.rm = TRUE),
    LastNonNA_Week = max(Week[!is.na(Cases)], na.rm = TRUE),
    PctMissing = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  filter(FirstNonNA_Week > 5) |>
  arrange(AN, DISTRICT, Disease) |>
  print(n = Inf)

###### Visualize the mid-year missingness patterns for the affected districts
# Step 1: identify the 2016 mid-year districts
midyear_2016_districts <- surveillance_doi |>
  filter(Disease != "Guinea Worm", AN == 2016) |>
  filter(!(DISTRICT == "Bekourou")) |>
  group_by(DISTRICT, Disease) |>
  summarise(
    FirstNonNA_Week = min(Week[!is.na(Cases)], na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(FirstNonNA_Week > 5) |>
  distinct(DISTRICT) |>
  pull(DISTRICT)

# Step 2: visualize week by week across diseases for these districts
surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT %in% midyear_2016_districts,
    Disease != "Guinea Worm"
  ) |>
  ggplot(aes(x = Week, y = Cases, color = Disease)) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 1, na.rm = TRUE) +
  facet_wrap(~ DISTRICT, scales = "free_y") +
  geom_vline(xintercept = 26, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 22, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 28, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 33, linetype = "dashed", color = "grey50") +
  labs(
    title = "2016 Mid-Year Districts: Weekly Cases by Disease",
    x = "Week",
    y = "Cases",
    color = "Disease"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
#pull first nonNA malaria week for each district in 2016, based on visual above malaria shows a lag, no other diseases reported
surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT %in% midyear_2016_districts,
    Disease == "Malaria"
  ) |>
  group_by(DISTRICT) |>
  summarise(
    FirstNonNA_Week_Malaria = min(Week[!is.na(Cases)], na.rm = TRUE),
    FirstCase_Week_Malaria = min(Week[!is.na(Cases) & Cases > 0], na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(FirstNonNA_Week_Malaria) |>
  print(n = Inf)
#plot Daffra because reporting seems to start at 18, but first case appears at week 37 
surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT == "Daffra",
    Disease == "Malaria"
  ) |>
  ggplot(aes(x = Week, y = Cases)) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  geom_vline(xintercept = 18, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 37, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 34, linetype = "dashed", color = "purple") +
  labs(
    title = "Daffra 2016 - Malaria Weekly Cases",
    subtitle = "Blue = Week 18 (first non-NA), Red = Week 37 (first case > 0)",
    x = "Week",
    y = "Cases"
  ) +
  theme_minimal()

# Clean up visualizations 
# Create object for first case week marker for cleaned up visualizations restricted to districts of interest based on missing patters above 
library(patchwork)
midyear_2016 <- c("Alifa", "Assinet", "Michemire", "Salal", "Daffra", "Youe")
#create first week marker object 
first_case_week <- surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT %in% midyear_2016,
    Disease == "Malaria"
  ) |>
  group_by(DISTRICT) |>
  summarise(
    FirstCase_Week_Malaria = min(Week[!is.na(Cases) & Cases > 0], na.rm = TRUE),
    .groups = "drop"
  )


first_case_week <- surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT %in% midyear_2016,
    Disease == "Malaria"
  ) |>
  group_by(DISTRICT) |>
  summarise(
    FirstCase_Week_Malaria = min(Week[!is.na(Cases) & Cases > 0], na.rm = TRUE),
    .groups = "drop"
  )


surveillance_doi |>
  filter(
    AN == 2016,
    DISTRICT %in% midyear_2016,
    Disease != "Guinea Worm"
  ) |>
  ggplot(aes(x = Week, y = Cases, color = Disease)) +
  geom_line(linewidth = 0.4, na.rm = TRUE) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_vline(
    data        = first_case_week,
    aes(xintercept = FirstCase_Week_Malaria),
    linetype    = "dashed",
    color       = "gray50",
    linewidth   = 0.5,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ DISTRICT, scales = "free_y", ncol = 2) +
  labs(
    title = "Figure S#: 2016 Mid-Year Activation Districts: Weekly Cases by Disease",
    x     = "Week",
    y     = "Cases",
    color = "Disease"
  ) +
  theme_classic() +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(color = "black", size = 8),
    legend.title     = element_text(color = "black", size = 8),
    plot.title       = element_text(color = "black", size = 8),
    axis.title       = element_text(color = "black", size = 8),
    axis.text        = element_text(color = "black", size = 8),
    axis.line        = element_line(size = 0.05, colour = "black"),
    axis.ticks       = element_line(size = 0.05, colour = "black"),
    strip.text       = element_text(color = "black", size = 8),
    strip.background = element_blank(), 
    panel.grid.major  = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor  = element_line(color = "gray95", linewidth = 0.2)
  )

ggsave(here("6_visuals", "S2_midyeardistricts.pdf"), width = 10, height = 8, dpi = 300)
```

STEP #4: Create clean dataset with placeholder districts removed and mid year activations adjusted for 
===========================================
Supplemental Figure: Second heat map for appendix that excludes placeholder and mid-year activations for missing analysis (placed here but the code that follows in the two chunks after is the diagnostic used to assess for district placeholders)
===========================================
```{r}
# Define full exclusions by year
full_exclusions_2015 <- c("Djamane Mbarissou", "Gagal", "Koyom", "Lagon", "Torrock", "Bekourou")
full_exclusions_2016_2017 <- c("Bekourou")

# Define full exclusion across years
surveillance_doi |> 
  filter(DISTRICT == "Isserom") |> distinct(AN) |> arrange(AN)
nonfunctional_exclusion <- c("Isserom")

# Define mid-year activation start weeks (for 2016)
midyear_activations_2016 <- tibble(
  DISTRICT = c("Alifa", "Assinet", "Michemire", "Salal", "Daffra"),
  StartWeek = c(33, 33, 33, 33, 37)
)

# Apply exclusions and mid-year activation filtering
surveillance_doi_clean <- surveillance_doi |>
  # Remove full placeholder exclusions by year
  filter(!(DISTRICT %in% full_exclusions_2015 & AN == 2015)) |>
  filter(!(DISTRICT %in% full_exclusions_2016_2017 & AN %in% c(2016, 2017))) |>
  # Remove non-functional site
  filter(!DISTRICT %in% nonfunctional_exclusion) |>
  # Apply mid-year activation cutoffs for 2016
  left_join(midyear_activations_2016, by = "DISTRICT") |>
  filter(
    is.na(StartWeek) |        # not a mid-year activation district, keep all
    AN != 2016 |              # not in 2016, keep all
    Week >= StartWeek         # in 2016, only keep from activation week onward
  ) |>
  select(-StartWeek)          # drop my helper column
```

STEP ## # Create heatmap with cleaned data
# Assessment of missingness patterns - excluding placeholder districts and mid-year activations
```{r}
missingness_heatmap_clean <- surveillance_doi_clean |>
  group_by(AN, Week, Disease) |>
  summarise(
    PctMissing = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  ggplot(aes(x = Week, y = AN, fill = PctMissing)) +
  geom_tile() +
  facet_wrap(~ Disease) +
  scale_fill_gradient(low = "white", high = "red", name = "% Missing") +
  scale_y_continuous(breaks = 2014:2025, expand = c(0, 0)) +
  labs(
    title = "S#: Missingness Heatmap by Week and Year Per Disease of Interest (Excl. Placeholder Districts & Mid-Year Activations)",
    x = "Week",
    y = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.text      = element_text(color = "black", size = 8),
    legend.title     = element_text(color = "black", size = 8),
    plot.title       = element_text(color = "black", size = 8),
    axis.title       = element_text(color = "black", size = 8),
    axis.text.y      = element_text(color = "black", size = 6),
    axis.text.x      = element_text(color = "black", size = 8),
    strip.text       = element_text(color = "black", size = 8),
    strip.background = element_blank()
  )

ggsave(here("6_visuals", "S2_missingness_heatmap_clean.pdf"), width = 10, height = 8, dpi = 300)
# Print to viewer

print(missingness_heatmap_clean)


```

STEP #4: Calculate missingness percentages with placeholder districts removed
===================================
Supplmental Table: Missingness Summary Table by Disease and Year with Placeholder Districts Removed by using surveillance_doi_clean dataset (for appendix)
===================================================
```{r}
#percent missing overall across all 12 years calculation excluding guinea worm
surveillance_doi_clean |>
  filter(Disease != "Guinea Worm") |>        # exclude Guinea Worm
  group_by(Disease, AN) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  select(Disease, AN, pct_miss) |>
  mutate(pct_miss = round(pct_miss, 1)) |>
  ungroup() |>
  summarize(avg_pct_miss = round(mean(pct_miss), 1))

#percent missing overall across all 12 years calculation including guinea worm 
surveillance_doi_clean |>
  group_by(Disease, AN) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  select(Disease, AN, pct_miss) |>
  mutate(pct_miss = round(pct_miss, 1)) |>
  ungroup() |>
  summarize(avg_pct_miss = round(mean(pct_miss), 1))
```

Additional investigation of missing in 2025 given it is higher than all prior years
```{r}

miss_var_summary(surveillance_doi_clean)

surveillance_doi_clean |>
  filter(AN == 2025) |>
  group_by(Disease) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  arrange(desc(pct_miss)) |>
  print(n = Inf)

# By district — is it specific reporting units dropping out?
surveillance_doi_clean |>
  filter(Disease != "Guinea Worm") |>
  filter(AN == 2025) |>
  group_by(DISTRICT) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  arrange(desc(pct_miss)) |>
  print(n = Inf)

# By week — is it system-wide gaps in particular weeks?
surveillance_doi_clean |>
  filter(Disease != "Guinea Worm") |>
  filter(AN == 2025) |>
  group_by(Week) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  arrange(desc(pct_miss)) |>
  print(n = Inf)

# By high frequency week above - is it disease specific in these weeks? 
surveillance_doi_clean |>
  filter(AN == 2025, Disease != "Guinea Worm", Week %in% c(27, 35)) |>
  group_by(Week, Disease) |>
  summarise(
    n_miss = sum(is.na(Cases)),
    n_total = n(),
    pct_miss = mean(is.na(Cases)) * 100,
    .groups = "drop"
  ) |>
  arrange(Week, desc(n_miss)) |>
  print(n = Inf)

```

STEP 4 Continued: appendix summary of missing values for all disease across years 

```{r}
miss_table <- surveillance_doi_clean |>
  group_by(Disease, AN) |>
  miss_var_summary() |>
  filter(variable == "Cases") |>
  ungroup() |>
  select(Disease, AN, pct_miss) |>
  mutate(pct_miss = round(pct_miss, 1)) |>
  pivot_wider(names_from = AN, values_from = pct_miss) |>
  arrange(Disease)

ft_miss <- flextable(miss_table) |>
  set_header_labels(Disease = "Disease") |>
  colformat_double(digits = 1) |>
  theme_booktabs() |>
  bold(part = "header") |>
  align(align = "center", part = "all") |>
  align(j = "Disease", align = "left", part = "all") |>
  fontsize(size = 8, part = "all") |>
  width(j = "Disease", width = 1.6) |>
  width(j = 2:ncol(miss_table), width = 0.45)
ft_miss
doc <- read_docx() |>
  body_add_par("Table S#. Percentage of Missing Cases by Disease and Year", style = "heading 2") |>
  body_add_flextable(ft_miss)

print(doc, target = here("6_visuals", "TableS_Missingness_by_Disease_Year.docx"))
```


